Project CENSE - WP5 Advancement Report

### 2019-03-21 report.tex
Final version

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{a4wide} % Smaller margins = more text per page.
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\pagestyle{fancy}
\begin{document}
\author{F\'elix GONTIER \\ \texttt{felix.gontier@ls2n.fr}}
\title{CENSE - Perceptual test II}
\maketitle

@@

This report presents the results of a perceptual experiment conducted as part of the Cense project~\footnote{\url{http://cense.ifsttar.fr/}}. From recordings of a sensor network with low information content, this study aims at proposing a processing chain for the estimation of pleasantness in urban sound environments.

\section{Test description}
\subsection{Objectives}
\label{sec:test_obj}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.6\textwidth]{figures/tvb_pres.eps}
    \caption{Estimated time of presence space for simulated scenes.}\label{fig:tvb_pres}
\end{figure}

@@

The experiment corpus totals 100 scenes.

@@

\subsection{Questions}

Test participants are asked to answer 9 questions per sound scene in the form of 11 point (0-10) scales. They are presented in french and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness (\textit{Unpleasant - Pleasant}) - P
\item Liveliness (\textit{Inert, amorphous - Lively, eventful}) - L
\item Overall loudness (\textit{Quiet - Noisy}) - OL
\item Interest (\textit{Boring, uninteresting - Stimulating, interesting}) - I
\item Calmness (\textit{Agitated, chaotic - Calm, peaceful}) - C
\end{itemize}
These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. While this work focuses on pleasantness, including other dimensions can be useful to verify the relevance of the data in relation to previous work.\\
Furthermore, the primary aim of this test is to model pleasantness from source-specific parameters. Previous studies show that pleasantness is mainly influenced by the activity of three sources: traffic, human voices and birds. The perception of source activity can be linked to notions of dominance, time of presence,  or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness can be different for background and emergent events. Thus, 4 additional questions are presented:
\begin{itemize}
\item Sound level of passing vehicles (\textit{Very low - Very high}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds (\textit{Never - Continuously}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
The corresponding interface is shown in Figure~\ref{fig:interface}.

@@

\subsection{Test conditions}

@@

A total of 23 students at Ecole Centrale de Nantes completed the test.

\section{Perceptual analysis}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=\textwidth]{figures/subj_gr.eps}
    \caption{Dendrogram of the hierarchical clustering of participants on the first 25 scenes of the test. No distinct groups are visible.}\label{fig:hclusters}
\end{figure}

\subsection{Correspondence between recorded and replicated scenes}

@@

Table~\ref{tab:ogrep} shows the mean differences between assessments for the recorded and corresponding replicated scenes. Kolmogorov-Smirnov statistical tests are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. This non parametric test is chosen because it assumes that distributions are continuous. Pleasantness is found to differ between the two groups only for one location, P4, which can be explained by high discrepancies in the perceptual time of presence of traffic and voices. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half
of its duration. To a lesser extent the same issue explains the large difference in assessed time of presence of voices in the location P3, but as voices have a less distinct influence on pleasantness this is not reflected directly. Differences can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness.\\

@@

\subsection{Perceptual space}

@@

Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace} as dots. Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace}a by black segments. The space covered by scenes based on new scenarios covers that of the studied real-life environments, demonstrating the efficiency of the scene generation procedure in terms of diversity.

@@

\subsection{Perceptual scales relationships}
\label{sec:percm}

\begin{table}
\centering
\caption{Pearson's correlation coefficients between perceptual scales at the scene level (n=100, *: p<0.05, **: p<0.01)}
\label{tab:percc}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P & 1 & -0.53** & -0.89** & 0.66** & 0.88** & -0.82** & -0.76** & 0.05 & 0.57** \\
	L &  & 1 & 0.76** & 0.06 & -0.78** & 0.39** & 0.17 & 0.60** & -0.36** \\
	OL &  &  & 1 & -0.39** & -0.96** & 0.71** & 0.59** & 0.17 & -0.45** \\
	I &  &  &  & 1 & 0.35** & -0.55** & -0.67** & 0.38** & 0.48** \\
	C &  &  &  &  & 1 & -0.67** & -0.55** & -0.24* & 0.48** \\
	$L_{T, p}$ &  &  &  &  &  & 1 & 0.79** & -0.15 & -0.41** \\
	$T_{T, p}$ &  &  &  &  &  &  & 1 & -0.35** & -0.42** \\
	$T_{V, p}$ &  &  &  &  &  &  &  & 1 & -0.21* \\
	$T_{B, p}$ &  &  &  &  &  &  &  &  & 1 \\ \hline
\end{tabular}
%}
\end{table}

@@

@@

\begin{align}
\hat P_{p1} &= 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p} + 0.14~T_{B, p}\\
\hat P_{p2} &= 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p} + 0.12~T_{B, p}
\end{align}

Both models have very similar formal expressions, and are comparable to the relation proposed in~\cite{ricciardi2014}, of expression (translated from 1-11 scales to 0-10):
\begin{equation}
\hat P_{p3} = 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}
\end{equation}.
Though, in the considered corpus the overall loudness and presence of voices have respectively a higher and lower impact on pleasantness. This is expected given the observed differences in correlations values.\\

Table~\ref{tab:percm} summarizes the performance of the three models. Again, the two models computed on the present corpus display similar, low estimation errors as well as no significant difference in adjusted $R^2$ and correlation to assessed pleasantness. In contrast, the model proposed in~\cite{ricciardi2014} results in higher prediction errors, along with less data variance explained by the model. However, this model was developed on a large corpus comprised of more than 3400 sound scenes and still shows acceptable performance for this new corpus: the mean standard deviation of pleasantness assessments for this experiment is about 1.77.

\begin{table}[t]
\centering
\caption{Performance of perceptual models for pleasantness prediction (**: p<0.01).}
\label{tab:percm}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c | c | c }
\hline
	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
	$\hat P_{p1}$ & 0.59 & 0.91 & 0.95** \\
	$\hat P_{p2}$ & 0.61 & 0.90 & 0.95** \\
	$\hat P_{p3}$~\cite{ricciardi2014} & 1.05 & 0.77 & 0.89** \\ \hline
\end{tabular}
%}
\end{table}

\section{Physical analysis}

@@

\subsection{Indicators}

@@

\begin{itemize}
\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level. The $L_{10}$ is associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
\item The time and frequency second derivative for the 4kHz band ($TFSD_{4kHz}$) introduced by~\cite{aumond2017}, that is found to correlate well with the perceptual time of presence of birds.
\end{itemize}

Furthermore, the simulation process outputs ground truth source contributions in the form of separate channels. Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:

@@

\begin{align}
\hat T_s(\alpha) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}_{\Delta_s(t)>\alpha}\\
\alpha_{opt} &= \argmax_{\alpha}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, T_s(\alpha)\right)
\end{align}

where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $t$ is the time frame and $T_{s, p}$ correspond to the perceptual time of presence assessments. While this indicator has a simple expression, it is not found to reflect well the properties of auditory masking resulting in a low optimal threshold $\alpha_{opt} = -19dB$.\\

@@

\begin{align}
\hat T_s(\alpha, \beta) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta_s(t, f)\mathbbm{1}_{\Delta_s(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta_s(t, f)>\alpha}}>\beta \right]\\
\alpha_{opt},\beta_{opt} &= \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, T_s(\alpha, \beta)\right)
\end{align}

The optimal threshold values are found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$. This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on bands where its level is at least $14dB$ below other sources.\\

@@

\subsection{Relation between perceptual an physical data}
\label{sec:physm}

@@

\begin{table}
\centering
\caption{Pearson's correlation coefficients between physical and perceptual indicators (n = 92, *: p<0.05, **: p<0.01). Non significant correlations at the 5\% threshold are noted NS.}
\label{tab:physc}
\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c | c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	$LA_{eq}$ & -0.86** & 0.68** & 0.92** & -0.37** & -0.88** & 0.77** & 0.66** & NS & -0.41** \\
	$LA_{50}$ & -0.84** & 0.67** & 0.91** & -0.33** & -0.87** & 0.71** & 0.63** & NS & -0.35** \\
	$L_{eq}$ & -0.88** & 0.67** & 0.91** & -0.44** & -0.88** & 0.83** & 0.71** & NS & -0.46** \\
	$L_{10}$ & -0.87** & 0.65** & 0.90** & -0.44** & -0.86** & 0.84** & 0.71** & NS & -0.47** \\
	$L_{50}$ & -0.89** & 0.65** & 0.92** & -0.43** & -0.89** & 0.77** & 0.71** & NS & -0.44** \\
	$L_{90}$ & -0.86** & 0.68** & 0.92** & -0.39** & -0.89** & 0.71** & 0.67** & NS & -0.40** \\
	$L_{50, 1kHz}$ & -0.88** & 0.69** & 0.92** & -0.42** & -0.89** & 0.74** & 0.73** & NS & -0.50** \\ \hline
	$TFSD_{4kHz}$ & 0.49** & -0.41** & -0.45** & 0.35** & 0.50** & -0.41** & -0.48** & -0.21* & 0.61** \\ \hline
	$L_{eq, T}$ & -0.58** & NS & 0.46** & -0.46** & -0.42** & 0.77** & 0.71** & NS & -0.36** \\
	$L_{eq, V}$ & NS & 0.50** & 0.31** & NS & -0.37** & NS & NS & 0.71** & -0.40** \\
	$L_{eq, B}$ & 0.27* & NS & NS & 0.35** & NS & -0.25* & -0.24* & NS & 0.71** \\ \hline
	$\Delta L_T$ & -0.45** & NS & 0.26* & -0.59** & -0.22* & 0.63** & 0.66** & -0.51** & -0.26* \\
	$\Delta L_V$ & NS & 0.50** & NS & 0.35** & NS & -0.27** & -0.38** & 0.59** & NS \\
	$\Delta L_B$ & 0.21* & -0.25* & -0.26* & NS & 0.25* & -0.24* & -0.25* & NS & NS \\ \hline
	$\hat T_T(\alpha, \beta)$ & -0.51** & NS & 0.33** & -0.56** & -0.27** & 0.58** & 0.82** & -0.36** & -0.33** \\
	$\hat T_V(\alpha, \beta)$ & NS & 0.44** & NS & 0.36** & NS & -0.23* & -0.41** & 0.83** & NS \\
	$\hat T_B(\alpha, \beta)$ & 0.58** & -0.33** & -0.47** & 0.55** & 0.52** & -0.45** & -0.56** & NS & 0.91** \\ \hline
	$\hat T_T(\alpha)$ & -0.47** & NS & 0.31** & -0.52** & -0.26* & 0.52** & 0.78** & -0.34** & -0.33** \\
	$\hat T_V(\alpha)$ & NS & 0.45** & NS & 0.30** & NS & NS & -0.30** & 0.81** & NS \\
	$\hat T_B(\alpha)$ & 0.61** & -0.37** & -0.50** & 0.57** & 0.56** & -0.46** & -0.56** & NS & 0.88** \\ \hline
\end{tabular}
}
\end{table}

Subsequently, multilinear models are constructed based on the results of the analysis of correlations. As in Section~\ref{sec:percm} a VIF check is performed on predictors to ensure that no multi-collinearity exists. The best pleasantness model optimized on the present data is:
\begin{equation}
\hat P_{\varphi 1} = 16.71 - 0.18~L_{50} + 1.01~\hat T_B(\alpha, \beta)
\end{equation}
The absence of indicators related to traffic and voices activity is explainable:
\begin{itemize}
\item The $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{p1}$ and $\hat P_{p2}$ respectively
\item In this corpus no strong contribution on pleasantness is identified from voices, with no significant correlation for both perceptual (Table~\ref{tab:percc}) and physical (Table~\ref{tab:physc}) indicators.
\end{itemize}
It is also possible to use the models discussed in Section~\ref{sec:percm} as is. In this situation, instead of predicting pleasantness directly, each intermediate perceptual parameter is modeled as a function of available physical variables. Here, there is already a one-to-one association between perceptual and physical quantities, and the models are linear:
\begin{align}
\hat{OL} &= 0.22~L_{50} - 8.4\\
\hat{T}_{T, p} &= 6.6~\hat T_{T}(\alpha, \beta) + 0.25\\
\hat{T}_{V, p} &= 6.3~\hat T_{V}(\alpha, \beta) + 1.54\\
\hat{T}_{B, p} &= 7~\hat T_{B}(\alpha, \beta) + 2.5
\end{align}
Substituting these equations in equation (3), the pleasantness model becomes:
\begin{equation}
\hat P_{\varphi 2} = 10.95 - 0.08~L_{50} - 0.99~\hat T_{T}(\alpha, \beta) + 1.26~\hat T_{V}(\alpha, \beta) + 1.05~\hat T_{B}(\alpha, \beta)
\end{equation}

Table~\ref{tab:physm} compares the performance of the two models. Both obtain good estimates of pleasantness for a 10-point scale. Again, the $\hat P_{\varphi 1}$ model optimized on this experiment's data is expectedly better, though less likely to generalize well to new scenes due to its construction process. It relies on the characteristics of the simulated scenes in terms of isolated samples variability and content, which leads to corpus-specific source contributions to pleasantness. This is particularly important in the case of voices, found in multiple previous studies to have a positive effect. Conversely, the $\hat P_{\varphi 2}$ model uses a perceptual model constructed on a much larger corpus of recorded scenes, and is less prone to these specificity properties.

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{a4wide} % Smaller margins = more text per page.
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\pagestyle{fancy}
\begin{document}
\author{F\'elix Gontier, Catherine Lavandier, Pierre Aumond, Mathieu Lagrange, Jean-Francois Petiot}
\title{Project CENSE - WP5 Advancement Report}
\maketitle

@@

This report presents the results of a perceptual experiment conducted as part of the Cense project~\footnote{\url{http://cense.ifsttar.fr/}}. From recordings of a sensor network with low information content, this study aims at proposing a processing chain for the estimation of pleasantness in urban sound environments using corpora of simulated scenes.\\

\section{Test description}
\label{sec:test}
\subsection{Objectives}
\label{sec:test_obj}

@@

\begin{figure}[h]
    \centering
     \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tv_pres.eps}
    \end{subfigure}%
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/tb_pres.eps}
    \end{subfigure}
    \begin{subfigure}[t]{0.33\textwidth}
        \centering
        \includegraphics[width=\textwidth]{figures/vb_pres.eps}
    \end{subfigure}
    \caption{Estimated time of presence of traffic, voices and birds sources for the corpus of simulated scenes with automatically generated scenarios.}\label{fig:tvb_pres}
\end{figure}

@@

The experiment corpus totals 100 scenes, including 6 recorded, 19 replicated and 75 simulated extracts.

@@

\subsection{Questions}
\label{sec:test_questions}

Test participants are asked to answer 9 questions per sound scene in the form of 11 point (0-10) scales. They are presented in french and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I
\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C
\end{itemize}
These quantities are typically studied in the perceptual characterization of sound scenes~\cite{axelsson2010, aumond2017, nilsson2007}. While this work focuses on pleasantness, including other dimensions can be useful to verify the relevance of the data in relation to previous work.\\
Furthermore, the primary aim of this test is to model pleasantness from source-specific parameters. Previous studies show that pleasantness is mainly influenced by the activity of three sources: traffic, human voices and birds. The perception of source activity can be linked to notions of dominance, time of presence,  or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness can be different for background and emergent events. Thus, 4 additional questions are presented:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
The corresponding interface is shown in Figure~\ref{fig:interface}.

@@

\subsection{Test conditions}
\label{sec:test_conditions}

@@

A total of 23 students aged 22 to 23 including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment.

\section{Perceptual analysis}
\label{sec:perc}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/subj_gr.eps}
    \caption{Dendrogram of the hierarchical clustering of participants on the first 25 scenes of the test. No distinct groups are visible.}\label{fig:hclusters}
\end{figure}

\subsection{Correspondence between recorded and replicated scenes}
\label{sec:perc_recrep}

@@

Table~\ref{tab:ogrep} shows the mean differences between assessments for the recorded and corresponding replicated scenes. Kolmogorov-Smirnov statistical tests are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. Pleasantness is found to differ between the two groups only for one location, P4, which can be explained by high discrepancies in the perceptual time of presence of traffic and voices. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half
of its duration. To a lesser extent the same issue explains the large difference in assessed time of presence of voices in the location P3, but as voices have a less distinct influence on pleasantness this is not reflected directly. Differences can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness.\\

@@

\subsection{Perceptual space}
\label{sec:perc_space}

@@

Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace} as dots. Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace}a by black arrows. The standard deviations of responses are shown as ellipses in Appendix~\ref{sec:app_a}, which overlap in most cases confirming the equivalence discussed in Section~\ref{sec:perc_recrep}. The space covered by scenes based on new scenarios (red dots in Figure~\ref{fig:pspace}b) covers that of the studied real-life environments (blue dots), demonstrating the efficiency of the scene generation procedure in terms of diversity.

@@

\subsection{Perceptual scales relationships}
\label{sec:perc_mdls}

@@

\begin{table}[t]
\centering
\caption{Pearson's correlation coefficients between perceptual scales at the scene level (n=100, *: p<0.05, **: p<0.01)}
\label{tab:percc}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P & 1 & -0.53** & -0.89** & 0.66** & 0.88** & -0.82** & -0.76** & 0.05 & 0.57** \\
	L &  & 1 & 0.76** & 0.06 & -0.78** & 0.39** & 0.17 & 0.60** & -0.36** \\
	OL &  &  & 1 & -0.39** & -0.96** & 0.71** & 0.59** & 0.17 & -0.45** \\
	I &  &  &  & 1 & 0.35** & -0.55** & -0.67** & 0.38** & 0.48** \\
	C &  &  &  &  & 1 & -0.67** & -0.55** & -0.24* & 0.48** \\
	$L_{T, p}$ &  &  &  &  &  & 1 & 0.79** & -0.15 & -0.41** \\
	$T_{T, p}$ &  &  &  &  &  &  & 1 & -0.35** & -0.42** \\
	$T_{V, p}$ &  &  &  &  &  &  &  & 1 & -0.21* \\
	$T_{B, p}$ &  &  &  &  &  &  &  &  & 1 \\ \hline
\end{tabular}
%}
\end{table}

@@

\begin{align}
\hat P_{p1} &= 8.56 - 0.63~OL - 0.20~L_{T, p} + 0.11~T_{V, p} + 0.14~T_{B, p}\\
\hat P_{p2} &= 8.99 - 0.67~OL - 0.15~T_{T, p} + 0.08~T_{V, p} + 0.12~T_{B, p}
\end{align}
Both models have very similar formal expressions, and are comparable to the relation proposed in~\cite{ricciardi2014}, of expression (translated from 1-11 scales to 0-10):
\begin{equation}
\hat P_{p3} = 7.11 - 0.38~OL - 0.15~T_{T, p} + 0.20~T_{V, p} + 0.15~T_{B, p}
\end{equation}
Though, in the considered corpus the overall loudness and presence of voices have respectively a higher and lower impact on pleasantness. This is expected given the observed differences in correlations values.\\

Table~\ref{tab:percm} summarizes the performance of the three models. Again, the two models computed on the present corpus display similar, low estimation errors as well as no significant difference in adjusted $R^2$ and correlation to assessed pleasantness. In contrast, the model proposed in~\cite{ricciardi2014} results in higher prediction errors, along with less data variance explained by the model. However, the error stays within acceptable range as it is below the mean standard deviation of pleasantness assessments of about 1.77 for this experiment.

\begin{table}[h]
\centering
\caption{Performance of perceptual models for pleasantness prediction (**: p<0.01).}
\label{tab:percm}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c | c | c }
\hline
	 & $RMSE$ & $R^2_{adj}$ & $r$ \\ \hline
	$\hat P_{p1}$ & 0.59 & 0.91 & 0.95** \\
	$\hat P_{p2}$ & 0.61 & 0.90 & 0.95** \\
	$\hat P_{p3}$~\cite{ricciardi2014} & 1.05 & 0.77 & 0.89** \\ \hline
\end{tabular}
%}
\end{table}

\section{Physical analysis}
\label{sec:phys}

@@

\subsection{Indicators}
\label{sec:phys_inds}

@@

\begin{itemize}
\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level. The $L_{10}$ is associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
\end{itemize}

The time and frequency second derivative for the 4kHz band with fast measurements ($TFSD_{4kHz, 1/8s}$) introduced by~\cite{aumond2017} is also computed as it represents well with the perceptual time of presence of birds. Furthermore, the simulation process outputs ground truth source contributions in the form of separate channels. Additional indicators are computed on the separated audio signals for the traffic, voices and birds sources:

@@

\begin{align}
\hat T_s(\alpha) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}_{\Delta_s(t)>\alpha}\\
\alpha_{opt} &= \argmax_{\alpha}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, T_s(\alpha)\right)
\end{align}
where $r$ is the Pearson's correlation coefficient, $s$ denotes the sound source, $t$ is the time frame and $T_{s, p}$ correspond to the perceptual time of presence assessments. While this indicator has a simple expression, it is not found to reflect well the properties of auditory masking resulting in a low optimal threshold $\alpha_{opt} = -19dB$.\\

@@

\begin{align}
\hat T_s(\alpha, \beta) &= \frac{1}{N_t}\sum_{t = 1}^{N_t}\mathbbm{1}\left[ \frac{\sum_{f = 1}^{N_f}\Delta_s(t, f)\mathbbm{1}_{\Delta_s(t, f)>\alpha}}{\sum_{f = 1}^{N_f}\mathbbm{1}_{\Delta_s(t, f)>\alpha}}>\beta \right]\\
\alpha_{opt},\beta_{opt} &= \argmax_{\alpha, \beta}\frac{1}{N_s}\sum_{s = 1}^{N_s}r\left(T_{s, p}, T_s(\alpha, \beta)\right)
\end{align}
The optimal threshold values are found as $\alpha_{opt} = -14dB$ and $\beta_{opt} = -7dB$. This means that to be considered heard in a given time frame, the source has to be emergent by more than $-7dB$ on average on bands where its level is at least $14dB$ below other sources.\\

@@

\subsection{Relation between perceptual an physical data}
\label{sec:phys_mdls}

@@

\begin{table}
\centering
\caption{Pearson's correlation coefficients between physical and perceptual indicators (n = 92, *: p<0.05, **: p<0.01). Non significant correlations at the 5\% threshold are noted NS.}
\label{tab:physc}
\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c | c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	$LA_{eq}$ & -0.86** & 0.68** & 0.92** & -0.37** & -0.88** & 0.77** & 0.66** & NS & -0.41** \\
	$LA_{50}$ & -0.84** & 0.67** & 0.91** & -0.33** & -0.87** & 0.71** & 0.63** & NS & -0.35** \\
	$L_{eq}$ & -0.88** & 0.67** & 0.91** & -0.44** & -0.88** & 0.83** & 0.71** & NS & -0.46** \\
	$L_{10}$ & -0.87** & 0.65** & 0.90** & -0.44** & -0.86** & 0.84** & 0.71** & NS & -0.47** \\
	$L_{50}$ & -0.89** & 0.65** & 0.92** & -0.43** & -0.89** & 0.77** & 0.71** & NS & -0.44** \\
	$L_{90}$ & -0.86** & 0.68** & 0.92** & -0.39** & -0.89** & 0.71** & 0.67** & NS & -0.40** \\
	$L_{50, 1kHz}$ & -0.88** & 0.69** & 0.92** & -0.42** & -0.89** & 0.74** & 0.73** & NS & -0.50** \\ \hline
	$TFSD_{4kHz, 1/8s}$ & 0.49** & -0.41** & -0.45** & 0.35** & 0.50** & -0.41** & -0.48** & -0.21* & 0.61** \\ \hline
	$L_{eq, T}$ & -0.58** & NS & 0.46** & -0.46** & -0.42** & 0.77** & 0.71** & NS & -0.36** \\
	$L_{eq, V}$ & NS & 0.50** & 0.31** & NS & -0.37** & NS & NS & 0.71** & -0.40** \\
	$L_{eq, B}$ & 0.27* & NS & NS & 0.35** & NS & -0.25* & -0.24* & NS & 0.71** \\ \hline
	$\Delta L_T$ & -0.45** & NS & 0.26* & -0.59** & -0.22* & 0.63** & 0.66** & -0.51** & -0.26* \\
	$\Delta L_V$ & NS & 0.50** & NS & 0.35** & NS & -0.27** & -0.38** & 0.59** & NS \\
	$\Delta L_B$ & 0.21* & -0.25* & -0.26* & NS & 0.25* & -0.24* & -0.25* & NS & NS \\ \hline
	$\hat T_T(\alpha, \beta)$ & -0.51** & NS & 0.33** & -0.56** & -0.27** & 0.58** & 0.82** & -0.36** & -0.33** \\
	$\hat T_V(\alpha, \beta)$ & NS & 0.44** & NS & 0.36** & NS & -0.23* & -0.41** & 0.83** & NS \\
	$\hat T_B(\alpha, \beta)$ & 0.58** & -0.33** & -0.47** & 0.55** & 0.52** & -0.45** & -0.56** & NS & 0.91** \\ \hline
	$\hat T_T(\alpha)$ & -0.47** & NS & 0.31** & -0.52** & -0.26* & 0.52** & 0.78** & -0.34** & -0.33** \\
	$\hat T_V(\alpha)$ & NS & 0.45** & NS & 0.30** & NS & NS & -0.30** & 0.81** & NS \\
	$\hat T_B(\alpha)$ & 0.61** & -0.37** & -0.50** & 0.57** & 0.56** & -0.46** & -0.56** & NS & 0.88** \\ \hline
\end{tabular}
}
\end{table}

Subsequently, multilinear models are constructed based on the results of the analysis of correlations. As in Section~\ref{sec:perc_mdls} a VIF check is performed on predictors to ensure that no multi-collinearity exists. The best pleasantness model optimized on the present data is:
\begin{equation}
\hat P_{\varphi 1} = 16.71 - 0.18~L_{50} + 1.01~\hat T_B(\alpha, \beta)
\end{equation}
The absence of indicators related to traffic and voices activity is explainable:
\begin{itemize}
\item The $L_{50}$ used as a predictor is correlated to both traffic parameters $L_{T, p}$ and $T_{T, p}$ that appeared in perceptual models $\hat P_{p1}$ and $\hat P_{p2}$ respectively
\item In this corpus no strong contribution on pleasantness is identified from voices, with no significant correlation for both perceptual (Table~\ref{tab:percc}) and physical (Table~\ref{tab:physc}) indicators.
\end{itemize}
It is also possible to use the models discussed in Section~\ref{sec:perc_mdls} as is. In this situation, instead of predicting pleasantness directly, each intermediate perceptual parameter is modeled as a function of available physical variables. Here, there is already a one-to-one association between perceptual and physical quantities, and the models are linear:
\begin{align}
\hat{OL} &= 0.22~L_{50} - 8.4\\
\hat{T}_{T, p} &= 6.6~\hat T_{T}(\alpha, \beta) + 0.25\\
\hat{T}_{V, p} &= 6.3~\hat T_{V}(\alpha, \beta) + 1.54\\
\hat{T}_{B, p} &= 7~\hat T_{B}(\alpha, \beta) + 2.5
\end{align}
Substituting these equations in equation (3), the pleasantness model becomes:
\begin{equation}
\hat P_{\varphi 2} = 10.95 - 0.08~L_{50} - 0.99~\hat T_{T}(\alpha, \beta) + 1.26~\hat T_{V}(\alpha, \beta) + 1.05~\hat T_{B}(\alpha, \beta)
\end{equation}
